Calendar-based graphics for visualizing people's daily schedules

### 2017-07-28 main.Rmd
minor tweaks

keywords:
  # at least one keyword must be supplied
  formatted: [data visualization, statistical graphics, time series, R package, grammar of graphics]
  plain:     [data visualization, statistical graphics, time series, R package, grammar of graphics]
preamble: >
  \usepackage{amsmath}
  \usepackage{tikz}
output: 
  bookdown::pdf_book:
    base_format: rticles::jss_article
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

We develop a method for organizing and visualizing temporal data, collected on sub-daily interval, into a calendar layout. We hope this calendar-based visualization to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area up until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of the foot traffic at 3 different sensors. Figure \ref{fig:facet-time} provides daily and sub-daily patterns by faceting on day of the week. The sensor data, like many temporal datasets on human behaviors, lends itself to a number of exploratory data visualization challenges:

1. Variations primarily result from multiple time scales including time of day, day of week, and day of year (such as public holiday and recurring events).
2. Since the data are often collected at sub-daily frequencies, they typically involve a large number of observations.
3. Measurements of a single type are made at multiple locations at a given time point, which creates the need for comparing and contrasting between locations.

@@

Alternatively, calendar-based graphics turn out to be a useful tool in unfolding human-related activities over time. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasons like school vacations; all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are limited to colour-encoding graphics and are unable to use time scales smaller than day. Time of day, which serves as one of the most important aspects in explaining variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using the curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

@@

(ref:flinders-polar-cap) Figure \ref{fig:flinders-2016} in circular layout, which is referred to as star plots. The periodic behaviors on workdays are clearly visible.

@@

The \code{free_wday} suggests that the scales are common for a given week day across all weeks, but different for each week days within a week. In other words, the same value within a given Monday, for example, corresponds to the same position across all Mondays but the different position in other week days. Similarly, the \code{free_mday} suggests free scaling for any day within a given month. The scaling option combined with the calendar layouts offers a number of varieties to construct the plot.

>>>>

keywords:
  # at least one keyword must be supplied
  formatted: [data visualization, statistical graphics, time series, R package, grammar of graphics]
  plain:     [data visualization, statistical graphics, time series, R package, grammar of graphics]
preamble: >
  \usepackage{amsmath}
output: 
  bookdown::pdf_book:
    base_format: rticles::jss_article
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

We develop a method for organizing and visualizing temporal data, collected on sub-daily interval, into a calendar layout. We hope this calendar-based visualization to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area up until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily rhythms, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of the foot traffic at 3 different sensors. Figure \ref{fig:facet-time} provides daily and sub-daily patterns by faceting on day of the week. The sensor data, like many temporal datasets on human behaviors, lends itself to a number of exploratory data visualization challenges:

1. Variations primarily source from multiple time scales including time of day, day of week, and day of year (such as public holiday and recurring events).
2. Since the data are often collected at sub-daily frequencies, they typically involve a large number of observations.
3. Measurements of a single type are made at multiple locations at a given time point, which creates the need for comparing and contrasting between locations.

@@

Alternatively, calendar-based graphics turn out to be a useful tool in unfolding human-related activities over time. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasons like school vacations; all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are limited to colour-encoding graphics and are unable to use time scales smaller than day. Time of day, which serves as one of the most important aspects in explaining substantial variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using the curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

@@

(ref:flinders-polar-cap) Figure \ref{fig:flinders-2016} in circular layout, which is referred to as star plots. The daily periodicity on work days are clearly visible.

@@

The \code{free_wday} suggests that the scales are common for a given week day across all weeks, but different for each week days within a week. In other words, the same value within a given Monday, for example, corresponds to the same position across all Mondays but the different position in other week days. Similarly, the \code{free_mday} suggests free scaling for any day within a given month. The scaling choices combined with the calendar layouts offers a number of varieties to construct the plot.

### 2017-07-28 main.tex
minor tweaks

\usepackage{amsmath} \usepackage{tikz}

@@

We develop a method for organizing and visualizing temporal data,
collected on sub-daily interval, into a calendar layout. We hope this
calendar-based visualization to provide insights into people's daily
schedules relative to events such as work days, weekends, holidays, and
special events. This work was originally motivated by studying foot
traffic in the city of Melbourne \citep{ped}. There have been 43 sensors
installed that count pedestrians every hour across the downtown area up
until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can
shed light into understanding people's daily schedules, or assisting
administration and business planning. We start off with the conventional
time series plot to catch a glimpse of such data. A small multiples,
shown in Figure \ref{fig:time-series-plot}, gives an overall picture of
the foot traffic at 3 different sensors. Figure \ref{fig:facet-time}
provides daily and sub-daily patterns by faceting on day of the week.
The sensor data, like many temporal datasets on human behaviors, lends
itself to a number of exploratory data visualization challenges:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Variations primarily result from multiple time scales including time
  of day, day of week, and day of year (such as public holiday and
  recurring events).
\item
  Since the data are often collected at sub-daily frequencies, they
  typically involve a large number of observations.
\item
  Measurements of a single type are made at multiple locations at a
  given time point, which creates the need for comparing and contrasting
  between locations.
\end{enumerate}

@@

Alternatively, calendar-based graphics turn out to be a useful tool in
unfolding human-related activities over time. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasons like school vacations; all of which have influence over
the turn-outs in the office. The calendar-based heatmap was implemented
in two R packages: \pkg{ggTimeSeries} \citep{R-ggTimeSeries} and
\pkg{ggcal} \citep{R-ggcal}. However, these techniques are limited to
colour-encoding graphics and are unable to use time scales smaller than
day. Time of day, which serves as one of the most important aspects in
explaining variations arisen from pedestrian sensor data, will be
neglected through daily aggregation. Additionally if simply using
colored blocks rather than curves, it may become perceptually difficult
to estimate the shape positions and changes, although using the curves
comes with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

@@

\caption[Figure \ref{fig:flinders-2016} in circular
layout, which is referred to as star plots. The periodic behaviors on
workdays are clearly visible.]{Figure \ref{fig:flinders-2016} in circular
layout, which is referred to as star plots. The periodic behaviors on
workdays are clearly visible.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}

>>>>

\usepackage{amsmath}

@@

We develop a method for organizing and visualizing temporal data,
collected on sub-daily interval, into a calendar layout. We hope this
calendar-based visualization to provide insights into people's daily
schedules relative to events such as work days, weekends, holidays, and
special events. This work was originally motivated by studying foot
traffic in the city of Melbourne \citep{ped}. There have been 43 sensors
installed that count pedestrians every hour across the downtown area up
until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can
shed light into understanding people's daily rhythms, or assisting
administration and business planning. We start off with the conventional
time series plot to catch a glimpse of such data. A small multiples,
shown in Figure \ref{fig:time-series-plot}, gives an overall picture of
the foot traffic at 3 different sensors. Figure \ref{fig:facet-time}
provides daily and sub-daily patterns by faceting on day of the week.
The sensor data, like many temporal datasets on human behaviors, lends
itself to a number of exploratory data visualization challenges:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Variations primarily source from multiple time scales including time
  of day, day of week, and day of year (such as public holiday and
  recurring events).
\item
  Since the data are often collected at sub-daily frequencies, they
  typically involve a large number of observations.
\item
  Measurements of a single type are made at multiple locations at a
  given time point, which creates the need for comparing and contrasting
  between locations.
\end{enumerate}

@@

Alternatively, calendar-based graphics turn out to be a useful tool in
unfolding human-related activities over time. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasons like school vacations; all of which have influence over
the turn-outs in the office. The calendar-based heatmap was implemented
in two R packages: \pkg{ggTimeSeries} \citep{R-ggTimeSeries} and
\pkg{ggcal} \citep{R-ggcal}. However, these techniques are limited to
colour-encoding graphics and are unable to use time scales smaller than
day. Time of day, which serves as one of the most important aspects in
explaining substantial variations arisen from pedestrian sensor data,
will be neglected through daily aggregation. Additionally if simply
using colored blocks rather than curves, it may become perceptually
difficult to estimate the shape positions and changes, although using
the curves comes with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

@@

\caption[Figure \ref{fig:flinders-2016} in circular
layout, which is referred to as star plots. The daily periodicity on
work days are clearly visible.]{Figure \ref{fig:flinders-2016} in circular
layout, which is referred to as star plots. The daily periodicity on
work days are clearly visible.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}
